---
title: "Case study: upper-quantile body-fat modelling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: upper-quantile body-fat modelling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5)
library(gkwqreg)
has_uqr <- requireNamespace("unitquantreg", quietly = TRUE)
knitr::opts_chunk$set(eval = has_uqr)
set.seed(20260827)
```

```{r noeval, eval = !has_uqr, echo = FALSE, results = "asis"}
cat("This vignette needs the `unitquantreg` package for its data. Install it to
run the code below.\n")
```

## The question

`bodyfat` (from `unitquantreg`, n = 298) records the android fat percentage of
participants along with age, sex and physical-activity level. Android fat — the
abdominal deposit — is the fraction clinically associated with cardiometabolic risk, and
the clinical interest is in the **upper tail**, not the average. A person at the 90th
percentile of android fat for their profile is the one worth flagging.

That is a quantile question, and it is the kind that mean regression answers only
indirectly.

```{r data}
data("bodyfat", package = "unitquantreg")
str(bodyfat[, c("android", "age", "sex", "ipaq")])
range(bodyfat$android)
```

The response lies strictly inside $(0,1)$, so the Generalized Kumaraswamy family applies
directly.

## Fitting the 0.9-quantile

```{r fit}
fit <- gkwqreg(android ~ age + ipaq + sex, data = bodyfat,
               tau = 0.9, family = "kw")
summary(fit)
```

Reading the quantile block:

- **`age`** is highly significant and positive. Under the logit link the coefficient
  is an effect on the log odds of the conditional 0.9-quantile.
- **`sexmale`** is strongly negative: at the same age and activity level, the upper tail of
  android fat sits lower for men in this sample.
- **`ipaq`** contributes little at this quantile.

The two summary lines worth reading before anything else:

```{r coverage}
c(empirical_coverage = mean(bodyfat$android <= fitted(fit)), target = 0.9)
```

The fit places 90.6% of observations below its fitted quantile against a target of 90%,
and the information-matrix condition number in the summary is small, so the standard
errors mean what they say.

## What the coefficients are worth on the response scale

```{r meff}
marginal_effects(fit)
```

One additional year of age shifts the conditional 0.9-quantile of android fat by the
amount in the `effect` column — directly interpretable, unlike the log-odds coefficient.

## Choosing a family: two criteria that disagree

```{r compare}
cmp <- suppressWarnings(
  compare_families(fit, families = c("kw", "ekw", "kkw", "bkw", "beta")))
cmp
```

This is worth dwelling on, because it is the situation the documentation warns about.
**AIC ranks `beta` first while the pinball loss ranks it last**, behind every Kumaraswamy member.
The two disagree because they measure different things: AIC scores the whole conditional
density, the pinball loss scores only the quantile being estimated. A family can describe the
bulk of the distribution well — which is what buys it the likelihood — and still place the
0.9-quantile worse than a rival that fits the bulk less well.

For a quantile question the pinball loss is the criterion that matches the goal — but it
must be computed **out of sample**, or it simply rewards flexibility:

```{r cv}
set.seed(1)
folds <- sample(rep(1:5, length.out = nrow(bodyfat)))
cv <- sapply(c("kw", "ekw", "bkw", "beta"), function(fam) {
  mean(sapply(1:5, function(k) {
    tr <- bodyfat[folds != k, ]; te <- bodyfat[folds == k, ]
    f <- suppressWarnings(try(gkwqreg(android ~ age + ipaq + sex, data = tr,
                                      tau = 0.9, family = fam), silent = TRUE))
    if (inherits(f, "try-error")) return(NA_real_)
    pinball(f, newdata = te)
  }), na.rm = TRUE)
})
round(cv, 6)
```

Five-fold cross-validated pinball loss is the number to act on.

## Where the model fits badly

```{r diag, fig.height = 6}
plot(fit, which = c(3, 4, 5, 6), nsim = 50)
```

Panel 5 is the one specific to quantile regression. It bins the `tau`-sign residuals
$1\{y_i \le \hat Q_i\} - \tau$ by fitted value. Under a correctly specified model each bin
mean is zero, and the bars are 95% binomial intervals — so a bin whose interval excludes
zero is direct, distribution-free evidence that the fit is miscalibrated in that part of
the covariate space.

## Across the distribution

The effect of age need not be constant across quantiles. That is the whole reason to fit
more than one.

```{r process, fig.height = 3.2}
fits <- suppressWarnings(
  gkwqreg(android ~ age + ipaq + sex, data = bodyfat,
          tau = c(0.1, 0.25, 0.5, 0.75, 0.9), family = "kw"))
qp <- quantile_process(fits)
plot(qp, parm = c("mu:age", "mu:sexmale"))
```

```{r crossing}
check_crossing(fits)
```

If the independently fitted levels cross anywhere, `rearrange()` restores monotonicity:

```{r rearr}
R <- rearrange(fits)
all(apply(R, 1, function(r) all(diff(r) >= 0)))
```

## Modelling the shape as well

Section 1.5 of the design notes recommends regressing the nuisance parameter on the same
covariates as the quantile whenever theory does not dictate an anchor. Doing so makes the
fit robust to the anchor choice, and here it also tests whether the *spread* of android fat
varies with age:

```{r het}
fit_het <- gkwqreg(android ~ age + ipaq + sex | age, data = bodyfat,
                   tau = 0.9, family = "kw")
anova(fit, fit_het)
```

```{r anchorcheck}
fa <- gkwqreg(android ~ age + ipaq + sex | age, data = bodyfat,
              tau = 0.9, family = "kw", anchor = "alpha")
rbind(beta_anchor  = coef(fit_het)[c("mu:age", "mu:sexmale")],
      alpha_anchor = coef(fa)[c("mu:age", "mu:sexmale")])
```

With the nuisance parameter saturated, the two anchors agree closely — which is exactly
the point of the recommendation: it removes the dependence on a choice you have no strong
grounds to make.

## Robust standard errors

If you doubt the distributional assumption but still want the parametric fit, the sandwich
estimator is available, and `sandwich`/`lmtest` work directly:

```{r sandwich}
round(cbind(model = sqrt(diag(vcov(fit))),
            sandwich = sqrt(diag(vcov(fit, type = "sandwich")))), 5)
```

A large gap between the two columns is itself a specification warning.

## Summary

The workflow this vignette follows is the one to reuse:

1. Fit at the quantile level the question is about.
2. Check empirical coverage and the condition number in `summary()`.
3. Choose the family by **out-of-sample pinball loss**, not AIC alone.
4. Read effects on the response scale with `marginal_effects()`.
5. Check calibration with panel 5, not just the Q-Q plot.
6. Saturate the nuisance parameters, and confirm the anchor no longer matters.
