---
title: "The anchor, and why it is a modeling choice"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The anchor, and why it is a modeling choice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gkwqreg)
set.seed(20260827)
```

## The reparametrization

For $0 < y < 1$ the Generalized Kumaraswamy density is

$$
f(y)=\frac{\lambda\alpha\beta}{B(\gamma,\delta+1)}\,
y^{\alpha-1}(1-y^{\alpha})^{\beta-1}
\bigl[1-(1-y^{\alpha})^{\beta}\bigr]^{\gamma\lambda-1}
\Bigl\{1-\bigl[1-(1-y^{\alpha})^{\beta}\bigr]^{\lambda}\Bigr\}^{\delta}
$$

with $\alpha,\beta,\gamma,\lambda>0$ and $\delta\ge0$. Unusually for a five-parameter
family, its quantile function is closed form. With
$z_\tau=\texttt{qbeta}(\tau,\gamma,\delta+1)$,

$$
Q(\tau)=\Bigl\{1-\bigl[1-z_\tau^{1/\lambda}\bigr]^{1/\beta}\Bigr\}^{1/\alpha}.
$$

```{r qcheck}
g <- expand.grid(t = c(.05, .5, .95), a = c(.5, 2), b = c(.7, 3),
                 gm = c(.8, 2), d = c(0, 2), L = c(.6, 2))
max(abs(gkwq_quantile(g$t, g$a, g$b, g$gm, g$d, g$L) -
        gkwdist::qgkw(g$t, g$a, g$b, g$gm, g$d, g$L)))
```

Now fix $\tau$, set $\mu = Q(\tau)$, and rearrange. The master identity is

$$\mu^{\alpha}=1-\bigl(1-z_\tau^{1/\lambda}\bigr)^{1/\beta},$$

and three of the five parameters can be solved for in closed form:

$$
\alpha=\frac{\log\bigl(1-(1-z_\tau^{1/\lambda})^{1/\beta}\bigr)}{\log\mu},
\quad
\beta=\frac{\log\bigl(1-z_\tau^{1/\lambda}\bigr)}{\log(1-\mu^{\alpha})},
\quad
\lambda=\frac{\log z_\tau}{\log\bigl(1-(1-\mu^{\alpha})^{\beta}\bigr)}.
$$

Each is a **ratio of two logarithms of quantities in $(0,1)$**, hence strictly positive
and finite for every $\mu,\tau\in(0,1)$ and every admissible value of the rest. That is
not a cosmetic property: it is why the optimizer needs no box constraints anywhere.

### It contains the prior art exactly

Set $\gamma=1$, $\delta=0$, $\lambda=1$. Then $z_\tau=\tau$ and the $\beta$-solve becomes

$$\beta=\frac{\log(1-\tau)}{\log(1-\mu^{\alpha})},$$

which is the Kumaraswamy quantile reparametrization of Mitnik and Baek (2013), and the
`kum` family of `unitquantreg`. The general formula *contains* the established result as
its two-parameter slice:

```{r mitnik}
anchor <- getFromNamespace(".gkwq_anchor_value", "gkwqreg")
mu <- 0.35; tau <- 0.25; a <- 1.8
c(general = anchor("beta", mu, tau, alpha = a, gamma = 1, delta = 0,
                   lambda = 1, z_mode = 1L),
  mitnik_baek = log1p(-tau) / log1p(-mu^a))
```

## Which parameter to anchor on

$\gamma$ and $\delta$ admit no closed-form solve: they enter only through
$z_\tau=I^{-1}_\tau(\gamma,\delta+1)$, and inverting the regularized incomplete beta in
its *shape* arguments has no elementary form. Every other family gets a closed form.

```{r anchors}
for (f in c("kw", "ekw", "kkw", "bkw", "gkw", "mc", "beta")) {
  cat(sprintf("%-5s anchors: %-22s parts: %s\n", f,
              paste(gkwq_anchors(f), collapse = ", "),
              paste(gkwq_parts(f), collapse = " | ")))
}
```

`mc` fixes $\alpha=\beta=1$, so neither is available and the $\lambda$-solve is forced.
`beta` fixes $\alpha=\beta=\lambda=1$, leaving only $\gamma,\delta$; there $\gamma$ is
solved numerically. The $\delta$-solve is inadmissible: $I_\mu(\gamma,\delta+1)$ ranges
only over $[\mu^\gamma, 1)$ as $\delta$ sweeps $[0,\infty)$, so a root exists only when
$\tau > \mu^{\gamma}$, which nothing guarantees. The $\gamma$-solve always has one.

## The finding that matters

Here is the point of this vignette. **The anchor looks like a reparametrization and is
not one, except under a specific condition.**

Generate data under the $\beta$-anchor with a covariate in the quantile, and fit the same
data three ways.

```{r experiment}
n <- 500
x <- runif(n, -2, 2)
mu <- plogis(0.4 + 1.1 * x)
y <- gkwdist::rkw(n, alpha = 2, beta = log1p(-0.5) / log1p(-mu^2))
d <- data.frame(y = y, x = x)

compare <- function(form) {
  fb <- gkwqreg(form, data = d, tau = 0.5, family = "kw", anchor = "beta")
  fa <- gkwqreg(form, data = d, tau = 0.5, family = "kw", anchor = "alpha")
  data.frame(
    specification = deparse(form),
    logLik_beta   = round(fb$loglik, 4),
    logLik_alpha  = round(fa$loglik, 4),
    gap           = signif(abs(fb$loglik - fa$loglik), 3),
    b1_beta       = round(fb$coef_list$mu[[2]], 3),
    b1_alpha      = round(fa$coef_list$mu[[2]], 3)
  )
}
```

```{r rows, echo = FALSE}
rows <- rbind(
  local({ fb <- gkwqreg(y ~ 1, data = d, tau = .5, family = "kw", anchor = "beta")
          fa <- gkwqreg(y ~ 1, data = d, tau = .5, family = "kw", anchor = "alpha")
          data.frame(specification = "y ~ 1  (nuisance constant)",
                     logLik_beta = round(fb$loglik, 4),
                     logLik_alpha = round(fa$loglik, 4),
                     gap = signif(abs(fb$loglik - fa$loglik), 3),
                     b1_beta = NA, b1_alpha = NA) }),
  compare(y ~ x),
  compare(y ~ x | x)
)
knitr::kable(rows, row.names = FALSE)
```

Read the `gap` column.

- **Row 1**: the quantile is intercept-only and the nuisance parameter is constant. The
  nuisance is *saturated relative to the quantile's predictor*, and the two anchors give
  the same likelihood to numerical precision. Here the anchor genuinely is a
  reparametrization.
- **Row 2**: the quantile varies with `x` while `alpha` is held constant. The likelihoods
  now differ by a large margin, and so does the coefficient of interest. These are
  **different models**, not different parametrizations of one model.
- **Row 3**: the nuisance is regressed on the same covariate. The gap collapses and the
  coefficient of interest agrees closely.

### The rule

The anchor is a pure reparametrization — a bijection with identical likelihood — **if and
only if the nuisance parameters are saturated relative to the predictor of the quantile.**

Anchoring on $\beta$ asserts "$\alpha$ constant, $\beta_i = f(\mu_i,\alpha)$". Anchoring
on $\alpha$ asserts the reverse. Those trace different paths through $(\alpha_i,\beta_i)$
space, and the data can tell them apart.

### What follows from it

Two anchors on the same data give **non-nested models of equal dimension**. A
likelihood-ratio test does not apply, and the package refuses to run one:

```{r refuse, error = TRUE}
fb <- gkwqreg(y ~ x, data = d, tau = 0.5, family = "kw", anchor = "beta")
fa <- gkwqreg(y ~ x, data = d, tau = 0.5, family = "kw", anchor = "alpha")
anova(fb, fa)
```

Use Vuong's test instead:

```{r vuong}
vuong_test(fb, fa)
```

**The practical recommendation**: unless theory dictates an anchor, regress the remaining
parameters on the same covariates as the quantile. Row 3 above is the evidence — it makes
the answer largely independent of a choice you have no strong grounds to make.

## Identifiability

Three separate facts, each enforced in code.

### `tau` is fixed, never estimated

The profile likelihood in $\tau$ is exactly flat: for any $\tau'$ there is a parameter
value reproducing the same distribution. $\tau$ indexes the *question*, not the model.

```{r tauflat}
f25 <- gkwqreg(y ~ 1, data = d, tau = 0.25, family = "kw")
f75 <- gkwqreg(y ~ 1, data = d, tau = 0.75, family = "kw")
c(tau_0.25 = f25$loglik, tau_0.75 = f75$loglik)
```

### $\delta=0$ confounds $\gamma$ and $\lambda$

When $\delta=0$, $I_u(\gamma,1)=u^{\gamma}$, so $F(y)=v^{\gamma\lambda}$ and only the
**product** is identified:

```{r confound}
yv <- c(.2, .5, .8)
max(abs(gkwdist::pgkw(yv, 1.5, 2, 3, 0, 0.7) -     # gamma*lambda = 2.1
        gkwdist::pgkw(yv, 1.5, 2, 0.7, 0, 3)))     # gamma*lambda = 2.1
```

The $\lambda$-anchor is therefore inadmissible whenever $\delta=0$ with $\gamma$ free,
because the quantile coefficients themselves become unidentified. The package errors
rather than warns. The default $\beta$-anchor is immune.

### The full `gkw` model is weakly identified

In *any* parametrization, with information-matrix condition numbers of order
$10^{8}$–$10^{11}$. `family = "gkw"` warns, and `summary()` reports the condition number
so you can judge for yourself. More parameters is not automatically better.

## A numerical note worth knowing about

Four of the seven families need $z_\tau=\texttt{qbeta}(\tau,\gamma,\delta+1)$ inside the
automatic-differentiation tape, differentiated with respect to its shape arguments. TMB
ships exactly that, and its shape derivatives are silently wrong: at
`TMB/include/tiny_ad/beta/toms708.cpp:341` the second shape is reduced by its integer
part and, when the remainder is exactly zero, a **literal** is assigned to an AD variable,
severing the derivative chain. The consequence is that `d/ddelta` comes back as exactly
`0` whenever $\delta+1$ is a whole number at or above a branch-dependent threshold — and
an optimizer reading a zero gradient leaves $\delta$ where it started and reports
convergence.

`gkwqreg` therefore ships its own incomplete-beta atomic
(`inst/include/gkwq_atomic.hpp`), with $\partial I/\partial p$ and $\partial I/\partial q$
from a series expansion. `vignette("gkwqreg-design")` sets out the reasoning and
demonstrates the corrected derivatives live; the unit test that pins them is `V10` in
`tests/testthat/test-gradient.R`.

Families `kw`, `ekw` and `kkw` never touch that code path at all: for them $z_\tau$ is
either $\tau$ itself or the elementary $1-(1-\tau)^{1/(\delta+1)}$.

## References

- Carrasco, Ferrari and Cordeiro (2010). A generalized Kumaraswamy distribution.
- Mitnik and Baek (2013). *Statistical Papers* **54**, 177–192.
- Cox and Reid (1987). Parameter orthogonality and approximate conditional inference.
- Boik and Robinson-Cox (1998). Derivatives of the incomplete beta function.
  *Journal of Statistical Software* **3**(1).
