---
title: "How gkwqreg relates to quantreg, unitquantreg and betareg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How gkwqreg relates to quantreg, unitquantreg and betareg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7)
library(gkwqreg)
set.seed(20260827)
has <- function(p) requireNamespace(p, quietly = TRUE)
```

## "Quantile" means three different things here

Most of the confusion in this corner of the ecosystem comes from one word carrying three
meanings.

**1. Check-function quantile regression.** `quantreg`, `qgam`. Distribution-free,
indexed by `tau`, no likelihood. Minimises the pinball loss directly.

**2. Location- or median-parametrized models.** `cdfquantreg`, and
`betareg`'s `predict(type = "quantile")`. A *single* fit whose location parameter happens
to be the median, or quantiles derived after the fact from a mean model. These are **not**
`tau`-indexed — `cdfquantreg` has no `tau` argument at all.

**3. `tau`-indexed parametric quantile regression.** `unitquantreg`, and this package.
One likelihood per `tau`, with a parameter that *is* the `tau`-quantile.

| package | support | type | families | nested? |
|---|---|---|---|---|
| `quantreg` | ℝ | check-function | — | — |
| `qgam` | ℝ | smooth check-function | — | — |
| `betareg` | (0,1) | mean–precision | beta | — |
| `cdfquantreg` | (0,1) | median / location | 36 pairs | no |
| `gamlss` | flexible | distributional | many (no Kumaraswamy) | no |
| `gkwreg` | (0,1) | distributional (mean) | 7 | **yes** |
| `unitquantreg` | (0,1) | **τ-indexed parametric** | 13 | no |
| **`gkwqreg`** | (0,1) | **τ-indexed parametric** | **7** | **yes** |

## The prior art, and the identity that proves it

The `kw` slice of this package **is** the Kumaraswamy quantile reparametrization of
Mitnik and Baek (2013), which `unitquantreg` implements as its `kum` family. That is not
a similarity to be glossed over — it is the strongest validation the construction has, so
the package tests it.

```{r uqr, eval = has("unitquantreg")}
library(unitquantreg)
n <- 400
x <- runif(n, -2, 2)
mu <- plogis(0.4 + 1.1 * x)
y <- gkwdist::rkw(n, alpha = 2, beta = log1p(-0.5) / log1p(-mu^2))
d <- data.frame(y = y, x = x)

out <- t(sapply(c(0.25, 0.50, 0.75), function(tau) {
  a <- gkwqreg(y ~ x, data = d, tau = tau, family = "kw")
  b <- unitquantreg(y ~ x, data = d, tau = tau, family = "kum")
  c(tau = tau,
    gkwqreg_b0 = coef(a)[[1]], gkwqreg_b1 = coef(a)[[2]],
    unitq_b0 = coef(b)[[1]],   unitq_b1 = coef(b)[[2]],
    max_diff = max(abs(unname(coef(a)[1:2]) - unname(coef(b)[1:2]))))
}))
knitr::kable(out, digits = c(2, 5, 5, 5, 5, 10))
```

On the author's machine the two agree to between `1.2e-08` and `4.4e-06`. They are the
same estimator, implemented independently — one by hand-derived C++ gradients, one by
automatic differentiation through a general five-parameter template.

**What is new here** is not the `kw` case. It is the extension to the four- and
five-parameter members — for which no quantile regression appears in print — and the fact
that they are **nested**, so family selection is an ordinary likelihood-ratio test.
`unitquantreg`'s thirteen families are mutually non-nested and require Vuong tests.

```{r nested}
d <- local({
  n <- 400; x <- runif(n, -2, 2); mu <- plogis(0.4 + 1.1 * x)
  data.frame(y = gkwdist::rkw(n, 2, log1p(-0.5) / log1p(-mu^2)), x = x)
})
anova(gkwqreg(y ~ x, data = d, tau = .5, family = "kw"),
      gkwqreg(y ~ x, data = d, tau = .5, family = "ekw"),
      gkwqreg(y ~ x, data = d, tau = .5, family = "gkw"))
```

## Against `quantreg`, under correct specification

```{r rq, eval = has("quantreg")}
library(quantreg)
loss <- function(y, q, tau) mean((y - q) * (tau - ((y - q) < 0)))
res <- t(sapply(c(0.25, 0.5, 0.75), function(tau) {
  a <- gkwqreg(y ~ x, data = d, tau = tau, family = "kw")
  r <- rq(y ~ x, data = d, tau = tau)
  c(tau = tau, gkwqreg = pinball(a), rq = loss(d$y, fitted(r), tau))
}))
knitr::kable(res, digits = 6)
```

The two are close, as they should be: with the family correctly specified, the parametric
fit has little to gain in sample and loses nothing. Its advantages lie elsewhere —
standard errors from a likelihood, nested model comparison, a full conditional
distribution, and no crossing.

## Against `quantreg`, under misspecification

This is where the honest comparison lives, and it must not be buried. Generate from a
two-component beta mixture, which is bimodal and lies outside the Generalized Kumaraswamy
family entirely.

```{r mis, eval = has("quantreg")}
n <- 600
x <- runif(n, -2, 2)
comp <- rbinom(n, 1, plogis(0.5 * x))
y <- ifelse(comp == 1, rbeta(n, 8, 2), rbeta(n, 2, 8))
y <- pmin(pmax(y, 1e-8), 1 - 1e-8)
dm <- data.frame(y = y, x = x)

## split-sample, so the comparison is out of sample
idx <- sample(n, n / 2)
tr <- dm[idx, ]; te <- dm[-idx, ]

mis <- t(sapply(c(0.1, 0.25, 0.5, 0.75, 0.9), function(tau) {
  a <- suppressWarnings(gkwqreg(y ~ x, data = tr, tau = tau, family = "ekw"))
  r <- rq(y ~ x, data = tr, tau = tau)
  c(tau = tau,
    gkwqreg = loss(te$y, predict(a, newdata = te), tau),
    rq      = loss(te$y, predict(r, newdata = te), tau))
}))
knitr::kable(mis, digits = 5)
```

Under a distribution the family cannot represent, check-function quantile regression
remains consistent for the conditional quantile and the parametric fit does not. That is
the central trade-off of parametric quantile regression, and it is the reason
`quantreg` remains the right default when you know nothing about the conditional
distribution.

The reverse holds too: when the family is right, the parametric fit uses information the
check function throws away, and it gives you a likelihood to do inference with.

## Against `betareg`

`betareg` fits a mean–precision model. Its `predict(type = "quantile")` applies the beta
quantile function to the fitted mean and precision, which is meaning (2) above.

```{r breg, eval = has("betareg")}
library(betareg)
data("GasolineYield", package = "betareg")

bq <- betareg(yield ~ batch + temp, data = GasolineYield)
gq <- gkwqreg(yield ~ batch + temp, data = GasolineYield, tau = 0.9, family = "kw")

## betareg: the 0.9-quantile implied by a MEAN model
q_breg <- qbeta(0.9, shape1 = predict(bq, type = "response") * predict(bq, type = "precision"),
                shape2 = (1 - predict(bq, type = "response")) * predict(bq, type = "precision"))

c(coverage_betareg = mean(GasolineYield$yield <= q_breg),
  coverage_gkwqreg = mean(GasolineYield$yield <= fitted(gq)),
  target = 0.9)
```

Both produce a number. Only one of them has a coefficient whose standard error refers to
the 0.9-quantile:

```{r bregcoef, eval = has("betareg")}
round(summary(gq)$coefficients["mu:temp", ], 5)
```

In `betareg` the coefficient on `temp` describes the **mean**. Converting it into a
statement about the 0.9-quantile means propagating uncertainty through both the mean and
the precision submodels; here the quantity is modelled directly.

## Choosing between them

- Conditional distribution unknown or clearly outside the family → **`quantreg`**.
- Smooth or additive quantile effects → **`qgam`**.
- Median only, no interest in shape → **`unitquantreg`** or **`cdfquantreg`**.
- A mean model that happens to also report quantiles → **`betareg`**, **`gkwreg`**.
- A `tau`-indexed likelihood with quantile-interpretable coefficients, shape
  heterogeneity, and nested family selection on (0,1) → **`gkwqreg`**.

## References

- Koenker and Bassett (1978). Regression quantiles. *Econometrica* **46**, 33–50.
- Mitnik and Baek (2013). *Statistical Papers* **54**, 177–192.
- Mazucheli, Alves, Menezes and Leiva (2022). `unitquantreg`. *CMPB* **221**, 106816.
- Ferrari and Cribari-Neto (2004). Beta regression for modelling rates and proportions.
- Chernozhukov, Fernández-Val and Galichon (2010). Quantile and probability curves
  without crossing.
