---
title: "Getting started with gkwqreg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with gkwqreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5)
library(gkwqreg)
set.seed(20260827)
```

## The question this package answers

Suppose a response lies strictly between 0 and 1 — a yield, a proportion, an index — and
you care about its upper tail rather than its average. Two families of tools exist.

**Check-function quantile regression** (`quantreg::rq`) is distribution-free and indexed
by a quantile level `tau`. It is the reference method and it stays consistent even when
you have no idea what the conditional distribution looks like.

**Distributional regression** (`gkwreg`, `betareg`, `gamlss`) fits a full parametric model
for the conditional distribution, and can report quantiles afterwards by applying the
quantile function to the fitted parameters. That answers *"given the fitted distribution,
what is its 90th percentile?"*

Neither answers *"how do covariates shift the 90th percentile?"* with a coefficient you
can read off and a standard error that refers to it. The first has no likelihood; the
second has no parameter that **is** the 90th percentile.

`gkwqreg` fixes `tau` in advance and reparametrizes the distribution so that one of its
parameters is exactly the conditional `tau`-quantile.

## A first fit

```{r data}
n  <- 400
x  <- runif(n, -2, 2)
z  <- rbinom(n, 1, 0.5)
mu <- plogis(0.4 + 1.1 * x - 0.6 * z)          # the conditional MEDIAN
y  <- gkwdist::rkw(n, alpha = 2, beta = log1p(-0.5) / log1p(-mu^2))
dat <- data.frame(y = y, x = x, z = factor(z))
```

The data are generated so that `plogis(0.4 + 1.1 x - 0.6 z)` is the conditional median.
Fitting at `tau = 0.5` should recover those numbers.

```{r fit}
fit <- gkwqreg(y ~ x + z, data = dat, tau = 0.5, family = "kw")
summary(fit)
```

Three things in that output are worth pausing on.

**The header** names the family, the quantile level and the anchored parameter. All three
change what the model *is*, so none of them is hidden.

**The coefficient block for `mu`** is labelled as effects on the log quantile odds, not on
a mean and not on the odds of an event. This is the most likely thing for a reader to get
wrong.

**Empirical coverage** reports the fraction of observations below the fitted quantile. It
should sit at `tau`, and it is the cheapest available check that the fit is doing what it
claims.

## The formula has one part per modelled parameter

Part one is always the conditional quantile. The remaining parts follow the family's
parameter order with the anchored parameter removed. `gkwq_parts()` is the contract:

```{r parts}
gkwq_parts("kw")
gkwq_parts("ekw")
gkwq_parts("gkw")
```

So for `kw` the second part is `alpha`:

```{r het}
fit_het <- gkwqreg(y ~ x + z | x, data = dat, tau = 0.5, family = "kw")
```

Omitted trailing parts default to `~ 1`. Supplying **more** parts than the family has is
an error that names the parts it expected, rather than silently ignoring the extras.

## What the coefficients mean

```{r meff}
marginal_effects(fit)
```

`exp(coef(fit)["mu:x"])` is the multiplicative effect on the odds of the conditional
median. `marginal_effects()` converts that to the effect on the median itself,
`beta_j * mu * (1 - mu)` under the logit link.

There is a small bonus in this parametrization. If a covariate appears in *both* the
quantile part and a nuisance part, the effect on the quantile is still exactly the
expression above, because `mu` **is** the quantile — the nuisance part changes the spread
of the conditional distribution, not the quantile being modelled. `marginal_effects()`
says so when it detects the overlap.

## Choosing a family

The seven families are genuinely nested, so a likelihood-ratio test applies:

```{r nesting}
fit_kw  <- gkwqreg(y ~ x + z, data = dat, tau = 0.5, family = "kw")
fit_ekw <- gkwqreg(y ~ x + z, data = dat, tau = 0.5, family = "ekw")
anova(fit_kw, fit_ekw)
```

For an out-of-sample choice, prefer the pinball loss: it is the criterion a quantile estimate
actually targets, whereas AIC compares likelihoods.

```{r compare}
compare_families(fit_kw, families = c("kw", "ekw", "bkw", "beta"))
```

## Diagnostics

```{r resid}
head(residuals(fit, type = "quantile"))
head(residuals(fit, type = "tau-sign"))
```

`"quantile"` residuals are standard normal under correct specification and are the
default. `"tau-sign"` is specific to quantile regression: under a correctly specified
model `E[1{Y <= Q_tau(X)} | X] = tau` exactly, so binning these by any covariate and
testing for a zero mean is a distribution-free calibration check with no analogue in mean
regression. Panel 5 of the diagnostic plot draws it.

```{r plot, fig.height = 6}
plot(fit, which = c(3, 4, 5, 6), nsim = 50)
```

## Several quantile levels

Each level is a separate likelihood, so a vector `tau` returns a container of independent
fits rather than one pooled object.

```{r process}
fits <- gkwqreg(y ~ x + z, data = dat, tau = c(0.1, 0.25, 0.5, 0.75, 0.9),
                family = "kw")
round(coef(fits), 3)
```

```{r procplot, fig.height = 3.2}
plot(quantile_process(fits), parts = "mu")
```

The coefficient path shows how the effect of `x` changes across the distribution — the
picture quantile regression exists to produce.

## Crossing

Independently fitted levels carry no guarantee that fitted quantiles increase with `tau`:

```{r crossing}
check_crossing(fits)
```

A *single* fit cannot cross, because its quantiles are those of one proper distribution
function:

```{r crossing1}
check_crossing(fit, taus = seq(0.1, 0.9, by = 0.1))
```

If separate fits do cross, `rearrange()` applies the monotone rearrangement of
Chernozhukov, Fernández-Val and Galichon (2010).

## Where to go next

The companion vignette, *The anchor, and why it is a modeling choice*, covers the
reparametrization itself, the identifiability constraints, and the one decision in this
package that changes your answers rather than your arithmetic.
