---
title: "Design notes: the vocabulary and the four decisions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design notes: the vocabulary and the four decisions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(gkwqreg)
set.seed(20260827)
```

Software is easier to use, and much easier to trust, when its words mean one thing and
its awkward choices are written down. This document is both halves: the vocabulary the
package commits to, and the four decisions a reader is most likely to look at and wonder
*why on earth did they do it that way?*

It is not a tutorial. For that, start with `vignette("gkwqreg")`; for the mathematics of
the reparametrization, `vignette("gkwqreg-anchor")`.

---

# Part I — The vocabulary

Terms are defined by what they **are**, not what they do. Where several words exist for
one concept, the package picks one and the alternatives are listed as *avoid* — not
because they are wrong in general, but because using two words for one thing is how
documentation starts lying.

## The quantile

**Quantile level.** The probability `tau` in `(0,1)` that indexes the question being
asked. It is supplied by the analyst, never estimated, and never enters the parameter
vector.
*Avoid*: quantile (on its own), percentile, alpha, p.

That `tau` is not estimable is a property of the model, not a convention. The profile
likelihood in `tau` is exactly flat: for any other level there is a parameter value
reproducing the same distribution, so the data cannot distinguish them.

```{r tauflat}
n <- 400
x <- runif(n, -2, 2)
mu <- plogis(0.4 + 1.1 * x)
y <- gkwdist::rkw(n, alpha = 2, beta = log1p(-0.5) / log1p(-mu^2))
d <- data.frame(y = y, x = x)

c(tau_0.25 = gkwqreg(y ~ 1, data = d, tau = 0.25, family = "kw")$loglik,
  tau_0.75 = gkwqreg(y ~ 1, data = d, tau = 0.75, family = "kw")$loglik)
```

Identical to numerical precision. `tau` indexes the *question*, not the model.

**Conditional quantile.** `mu_tau(x)`, the value the response falls below with
probability `tau` given covariates `x`. It is the regression target and the value
`fitted()` returns.
*Avoid*: mu (on its own — it collides with the mean), location, fitted mean, prediction.

```{r fitted}
fit <- gkwqreg(y ~ x, data = d, tau = 0.9, family = "kw")
c(coverage = mean(fit$y <= fitted(fit)), target = 0.9)
```

**Quantile odds.** `mu_tau / (1 - mu_tau)`. Under the default logit link, exponentiating a
coefficient of the conditional quantile gives a multiplicative effect on *this* quantity —
not on the odds of an event, and not on a mean.
*Avoid*: odds ratio, relative risk.

**Pinball loss.** Mean of `(y - q)(tau - 1{y < q})` over the sample. The loss a quantile
estimate actually targets, and the criterion for comparing families out of sample.
*Avoid*: check loss, quantile loss, tick loss.

```{r pinball}
c(pinball = pinball(fit),
  from_residuals = mean(residuals(fit, type = "check")))
```

## The reparametrization

**Anchor.** The parameter eliminated from the family and replaced by the conditional
quantile. A modeling choice, not an internal detail: two anchors on the same data are
different models of equal dimension unless the nuisance parameters are saturated.
*Avoid*: pivot, eliminated parameter, reparametrization target, solve-for.

**Anchor solve.** The expression giving the anchor from the conditional quantile, the
quantile level and the nuisance parameters. For six of the seven families it is closed
form — a ratio of two negative logarithms, hence strictly positive without any box
constraint. The `beta` family is the exception: its anchor is `gamma`, the equation
`I_mu(gamma, delta + 1) = tau` does not invert, and `gamma` comes from a one-dimensional
root find.
*Avoid*: inversion, back-transform, link.

That last clause is the load-bearing one. For the Kumaraswamy the solve is
`beta = log(1 - tau) / log(1 - mu^alpha)`: both logarithms are negative for any
`mu, tau` in `(0,1)`, so their ratio is positive by construction and the optimizer needs
no constraint anywhere.

**Nuisance parameter.** Any family parameter that is neither the anchor nor the
conditional quantile. Modelable in its own right; *nuisance* names its role in this
parametrization, not its importance.
*Avoid*: shape parameter, dispersion, precision, ancillary.

**Saturated.** Said of the nuisance parameters when each is regressed on at least the
covariates used for the conditional quantile. Under saturation, and only then, the anchor
choice is a pure reparametrization with identical likelihood.
*Avoid*: full model, unrestricted, fully specified.

## The family

**Family.** One of the seven members of the Generalized Kumaraswamy lattice: `gkw`, `bkw`,
`kkw`, `ekw`, `mc`, `kw`, `beta`. Members are genuinely nested, so likelihood-ratio tests
compare them.
*Avoid*: distribution, model, link family, sub-model.

The nesting is a real containment, not a family resemblance, so a larger family can never
fit worse **at its maximum**. The lattice is not a single chain, so the comparison has to
be made along one:

```{r nesting}
ll <- vapply(c("kw", "ekw", "kkw", "bkw", "gkw"), function(f) {
  suppressWarnings(gkwqreg(y ~ x, data = d, tau = 0.5, family = f))$loglik
}, numeric(1))
round(ll[c("kw", "ekw", "kkw", "gkw")], 4)   # one chain
round(ll[c("kw", "bkw", "gkw")], 4)          # the other
```

What is printed are optimizer outputs, not maxima, and the distinction is visible here:
`gkw` lands below `kkw`, which containment forbids. That is not evidence against the
lattice — it is the diagnostic that the five-parameter fit stopped short, which is
exactly what the package warns about when it reports `gkw` as weakly identified. Read a
violation of monotonicity as a message about the optimizer, never about the data.

**Part.** One component of the multi-part formula, naming a parameter that carries its own
linear predictor. Part one is always the conditional quantile; the rest follow family
order with the anchor removed.
*Avoid*: component, submodel, equation, block.

```{r parts}
for (f in c("kw", "ekw", "gkw", "mc", "beta")) {
  cat(sprintf("%-5s %s\n", f, paste(gkwq_parts(f), collapse = " | ")))
}
```

## Across quantile levels

**Quantile process.** The collection of fits over a grid of quantile levels, and the
coefficient paths it traces. Each level is a separate likelihood, so each fit keeps its own
log-likelihood and covariance.
*Avoid*: multi-tau model, quantile curve, joint fit.

The package enforces this rather than trusting the reader to remember it:

```{r container, error = TRUE}
fits <- gkwqreg(y ~ x, data = d, tau = c(0.25, 0.5, 0.75), family = "kw")
logLik(fits)
```

**Crossing.** Fitted conditional quantiles that fail to increase with the quantile level.
Possible across independent fits; impossible within one fit, whose quantiles are those of a
proper distribution function.
*Avoid*: monotonicity violation, inversion, non-monotone quantiles.

---

# Part II — The four decisions

Each of these was a genuine fork with a defensible alternative. They are recorded because
a future reader will otherwise reasonably assume the opposite was intended.

## Decision 1 — The anchor is a public modeling argument

Eliminating a parameter in favour of the conditional quantile looks like a
reparametrization, and under saturated nuisance parameters it is one: the likelihood is
identical whichever parameter is anchored. It stops being one the moment the conditional
quantile varies with covariates while a nuisance parameter is held homogeneous.

Anchoring on `beta` asserts *"`alpha` constant, `beta_i = f(mu_i, alpha)`"*. Anchoring on
`alpha` asserts the reverse. Those trace different paths through parameter space, and the
data can tell them apart:

```{r anchor}
rows <- lapply(list(`y ~ 1     (nuisance constant)`  = y ~ 1,
                    `y ~ x     (nuisance constant)`  = y ~ x,
                    `y ~ x | x (nuisance saturated)` = y ~ x | x),
  function(f) {
    b <- gkwqreg(f, data = d, tau = 0.5, family = "kw", anchor = "beta")
    a <- gkwqreg(f, data = d, tau = 0.5, family = "kw", anchor = "alpha")
    data.frame(logLik_beta = round(b$loglik, 3),
               logLik_alpha = round(a$loglik, 3),
               gap = signif(abs(b$loglik - a$loglik), 3))
  })
knitr::kable(do.call(rbind, rows))
```

Read the `gap` column. It is zero in the first row, where the quantile has no covariates
and the nuisance is trivially saturated. It is enormous in the second, where the quantile
varies with `x` while `alpha` is held constant. It collapses again in the third, once
`alpha` is regressed on the same covariate — not to exactly zero, because the two fits
still stop at slightly different points on a very flat surface, but to a difference of no
practical consequence.

**The rule.** The anchor is a pure reparametrization — a bijection with identical
likelihood — *if and only if* the nuisance parameters are saturated relative to the
predictor of the conditional quantile.

**Consequences the package enforces.** `anchor` is an argument of `gkwqreg()`, defaulted
per family and printed by `summary()`. Two anchors give non-nested models of *equal
dimension*, so a likelihood-ratio test does not apply and `anova()` refuses to run one:

```{r anovarefuse, error = TRUE}
fb <- gkwqreg(y ~ x, data = d, tau = 0.5, family = "kw", anchor = "beta")
fa <- gkwqreg(y ~ x, data = d, tau = 0.5, family = "kw", anchor = "alpha")
anova(fb, fa)
```

`vuong_test(fb, fa)` is the right tool. And the practical advice, which the third row
above is the evidence for: regress the nuisance parameters on the same covariates as the
quantile, and the answer stops depending on a choice you have no grounds to make.

**The alternative we rejected** was hiding the anchor as an implementation detail and
choosing it for the user. That would have been defensible if it were a reparametrization.
It is not.

## Decision 2 — The package ships its own incomplete-beta atomic

Four of the seven families need `z_tau = qbeta(tau, gamma, delta + 1)` inside the
automatic-differentiation tape, differentiated with respect to its **shape** arguments.
TMB ships exactly that, with reverse-mode derivatives via the implicit function theorem.
It is silently wrong.

At `TMB/include/tiny_ad/beta/toms708.cpp:337-342` the second shape argument is reduced by
its integer part:

```c
n = (int) trunc(b0);
b0 -= n;
if (b0 == 0.) { --n; b0 = 1.; }
```

That last statement assigns a **literal** to an AD variable, severing the derivative
chain. TMB's `qbeta` atomic obtains its shape partials from that same `pbeta`, so it
inherits the defect. Reproduced against `numDeriv` at `p = 0.37` during development:
`d/ddelta` comes back as **exactly 0** whenever `delta + 1` is a whole number at or above a
branch-dependent threshold — 2 for `gamma = 3`, 3 for `gamma = 1.6` — while `d/dgamma` is
correct throughout.

The consequence is not a crash. An optimizer reading a zero gradient concludes it is at an
optimum in that coordinate, leaves `delta` at its starting value, and reports convergence.
The fit looks clean and is wrong.

So `inst/include/gkwq_atomic.hpp` supplies `dI/dp` and `dI/dq` from a series expansion of
the regularized incomplete beta, with the standard reflection rule and a branch-free
digamma, wrapped in two `TMB_ATOMIC_STATIC_FUNCTION` atomics. Validated against `numDeriv`
over 720 shape and argument combinations: maximum error `7.5e-11` on the derivatives,
`7.4e-15` on the value, median 17 iterations.

Here is the check, run live, through the public interface. A fitted `bkw` model carries
its TMB object, whose gradient can be evaluated at **any** parameter value. The fit itself
is incidental here — it exists only to build the tape. The derivative is then evaluated at
a fixed, well-conditioned point of our choosing, with `delta + 1` placed exactly on the
whole numbers that break TMB's own version:

```{r atomic}
dg <- data.frame(y = gkwdist::rgkw(300, 1.5, 2.0, 1.3, 1.0, 1.0),
                 x = runif(300, -1, 1))
fb <- suppressWarnings(gkwqreg(y ~ x, data = dg, tau = 0.5, family = "bkw"))

## A fixed evaluation point, NOT wherever the optimizer stopped. Evaluating at
## an estimate would make the demonstration depend on the optimizer, and a
## degenerate fit sits inside the numerical guards where a finite difference and
## an exact derivative legitimately disagree.
p <- c("mu:(Intercept)"    = 0,          "mu:x"              = 0.5,
       "alpha:(Intercept)" = log(1.5),   "gamma:(Intercept)" = log(1.3),
       "delta:(Intercept)" = 0)
i <- match("delta:(Intercept)", names(p))

out <- t(vapply(c(1, 2, 3, 4), function(delta) {
  p[[i]] <- log(delta)                    # delta + 1 = 2, 3, 4, 5 exactly
  ad <- as.numeric(fb$obj$gr(p))
  nd <- numDeriv::grad(fb$obj$fn, p)
  c(`delta+1` = delta + 1, AD = ad[i], numDeriv = nd[i],
    rel_err = max(abs(ad - nd)) / max(1, max(abs(nd))))
}, numeric(4)))
knitr::kable(out, digits = c(0, 6, 6, 10))
```

The `AD` column is non-zero and agrees with `numDeriv` to nine or ten digits. With TMB's
built-in `qbeta` supplying the shape partials it would read `0.000000` in every row, and
nothing would announce it.

**The alternatives.** Nudging whole-number shapes by `1e-8` in the starting values is
verified to restore correct derivatives and is far less code — but it protects only the
starting point, and nothing stops the optimizer from landing on a whole number later, where
the failure is silent again. Boik and Robison-Cox (1998) is the reference algorithm and
additionally yields second derivatives, but it is an order of magnitude more code and
Decision 4 means second-order AD is never needed. The series is the smallest thing that is
actually correct everywhere.

**Families `kw`, `ekw` and `kkw` never touch this code path.** For them `z_tau` is either
`tau` itself or the elementary `1 - (1-tau)^(1/(delta+1))`, which is why those three carry
no differentiation risk at all.

## Decision 3 — Depend on `gkwdist`, suggest `gkwreg`, and do not inherit

`gkwreg` fits the same seven families and owns roughly 400 to 600 lines of formula, link
and model-frame plumbing that this package would otherwise rewrite. Every line of it is a
dot-prefixed internal, so reusing it means either `gkwreg:::` — a check NOTE that CRAN
rejects for new submissions — or growing `gkwreg`'s public API with functions that exist
only to serve a sibling. `gkwdist` is already the shared distribution kernel, carries no
regression opinions, and keeps the dependency graph acyclic.

Class `gkwqreg` therefore does **not** inherit from `gkwreg`. Inheritance would make every
missing method silently dispatch mean semantics on a quantile object, and the most
dangerous case is `fitted()`:

```{r fittedsem}
c(quantile_0.9 = mean(fitted(fit)),
  mean         = mean(fitted(fit, type = "mean")))
```

Those are different quantities. A missing method must error, not guess.

**The cost, stated plainly.** The plumbing is reimplemented rather than shared, so the two
packages can drift. That is accepted: cross-validation tests under
`skip_if_not_installed("gkwreg")` catch drift where it matters, and promoting the shared
layer into `gkwdist` remains open as a later move.

## Decision 4 — One template compiled at install time, and `optimHess` for observed information

`gkwreg` ships seven TMB templates and compiles them at **run** time. That needs a C++
toolchain on the user's machine — which CRAN Windows and macOS binary users do not have —
and costs 40 to 90 seconds per family inside `R CMD check`.

This package ships a single generic template taking `DATA_INTEGER` codes for the family and
the anchor, compiled into `src/` at install time. A family branch is free at run time
because TMB tapes only the executed path, and one translation unit keeps install time
within CRAN's budget. Dropping Rcpp removes `RcppArmadillo` from `LinkingTo` and the
`R_init_gkwqreg` / `TMB_LIB_INIT` double-registration clash.

Nothing in the template switches on a family *name*: the family arrives entirely as data,
so `R/gkwqreg-families.R` is the single source of truth and adding a family costs a
registry entry rather than a new template.

The second half of this decision is about **observed information**. It comes from
`optimHess` applied to the automatic-differentiation gradient, which is what `sdreport()`
already does for models without random effects. Two tempting alternatives are both wrong
here:

- `obj$he()` would require second-order differentiation *through* the incomplete-beta
  atomic, for `bkw`, `gkw`, `mc` and `beta` — the same four named above.
- The naive `J' H J` sandwich of the unreparametrized Hessian omits the curvature term
  `sum_k g_k grad^2 theta_k` and is simply wrong in a regression.

Because the conditional quantile is **not** orthogonal to the remaining parameters in the
Cox-Reid sense, the off-diagonal blocks matter and the full matrix is what gets reported.
`summary()` prints the information matrix's condition number so a weakly identified fit is
visible rather than silent:

```{r cond}
vapply(c("kw", "ekw", "gkw"), function(f) {
  suppressWarnings(gkwqreg(y ~ x, data = d, tau = 0.5, family = f))$cond_number
}, numeric(1))
```

---

# What these decisions cost

Being explicit about the wins is easy; the trade-offs are the part worth writing down.

| Decision | What it costs |
|---|---|
| Anchor is public | One more argument users must think about, and a default that can be wrong for their data |
| Own atomic | ~250 lines of numerical code this package must maintain and keep tested |
| No inheritance from `gkwreg` | Plumbing duplicated across two packages, which can drift |
| One precompiled template | Install-time compilation, and a template driven by data rather than readable per-family code |

Each is a deliberate trade, and the tests that hold them in place are `V6`/`V7` for the
anchor, `V10` for the atomic, the `fitted()` semantics tests for inheritance, and the
nesting-lattice test for the shared template.

# References

- Boik, R. J. and Robison-Cox, J. F. (1998). Derivatives of the incomplete beta function.
  *Journal of Statistical Software* **3**(1), 1-20. [doi:10.18637/jss.v003.i01](https://doi.org/10.18637/jss.v003.i01)
  (The journal's own article page misspells the second author as "Robinson-Cox".
  "Robison-Cox" is the spelling in Crossref's deposit and on the author's own
  faculty page at Montana State University; it is the one used here.)
- Cox, D. R. and Reid, N. (1987). Parameter orthogonality and approximate conditional
  inference. *JRSS-B* **49**(1), 1-18.
- Mitnik, P. A. and Baek, S. (2013). The Kumaraswamy distribution: median-dispersion
  re-parameterizations for regression modeling and simulation-based estimation.
  *Statistical Papers* **54**(1), 177-192. [doi:10.1007/s00362-011-0417-y](https://doi.org/10.1007/s00362-011-0417-y)
